---
title: "eda project - WNBA"
author: "Erin Franke"
date: "6/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Load the data 
```{r}
library(wehoop)
library(tidyverse)
library(stringr)
library(ggthemes)
library(ggtext)
library(sportyR)
#devtools::install_github(repo = "saiemgilani/wehoop", force = TRUE)
wnba_pbp_data <- load_wnba_pbp(2021)

# Get the shots and clean this data a bit:
wnba_shots_data <- wnba_pbp_data %>%
  filter(shooting_play)
# Remove unnecessary columns:
wnba_shots_data <- wnba_shots_data %>%
  dplyr::select(-shooting_play, -id, -participants_2_athlete_id,
                -type_abbreviation, -season, -season_type, 
                -home_team_spread, -game_spread, -home_favorite)
```

Basic data exploration
```{r}
#find number of games in the dataset
wnba_shots_data %>%
  group_by(game_id) %>%
  count() %>%
  nrow()

#find how many away games each team played
away_games <- wnba_shots_data %>%
  group_by(away_team_name) %>%
  count(game_id) %>%
  group_by(away_team_name) %>%
  count() %>%
  rename(n_away = n)
home_games <- wnba_shots_data %>%
  group_by(home_team_name) %>%
  count(game_id) %>%
  group_by(home_team_name) %>%
  count() %>%
  rename(n_home = n)
home_games %>%
  full_join(away_games, by = c("home_team_name" = "away_team_name")) %>%
  mutate(total_games = n_home + n_away) %>%
  arrange(desc(total_games))
```

### Visualization 1

Filter the data for all shots in the 4th quarter and overtime where the score is within one possession. 
```{r}
clutchshots <- wnba_shots_data %>%
  filter(period_display_value %in% c("4th Quarter", "OT"), abs(home_score - away_score) <= 3) %>%
  mutate(score_difference = abs(home_score - away_score),
         shooter = word(text, 1, 2)) %>%
  select(game_id, home_score, away_score, score_difference, shooter, scoring_play, clock_display_value, type_text, text, clock_minutes, clock_seconds, everything()) 

#filter for players with at least 15 shots in these clutch situations
quarter4players <- clutchshots %>%
  group_by(shooter) %>%
  count() %>%
  arrange(desc(n)) %>%
  head(n = 11) 

#use an inner join to filter the main dataset and calculate the clutch shooting percentage for each player
(clutch_players <- clutchshots %>%
  inner_join(quarter4players, by = "shooter") %>%
  group_by(shooter, scoring_play) %>%
  summarize(count = n()) %>%
  pivot_wider(names_from = scoring_play, values_from = count) %>%
  rename(`shot missed` = `FALSE`, `shot made` = `TRUE`) %>%
  mutate(`clutch shot percentage` = `shot made` /(`shot made`+`shot missed`)) %>%
  arrange(desc(`clutch shot percentage`)))
```

Find these players shooting percentages in "non-clutch" situations. 
```{r}
`%!in%` <- Negate(`%in%`)
(non_clutch_situations <- wnba_shots_data %>%
  filter(period_display_value %!in% c("4th Quarter", "OT")| abs(home_score - away_score) > 3) %>%
  mutate(score_difference = abs(home_score - away_score),
         shooter = word(text, 1, 2)) %>%
  select(game_id, home_score, away_score, score_difference, shooter, scoring_play, clock_display_value, type_text, text, clock_minutes, clock_seconds, everything()) %>%
  inner_join(quarter4players, by = "shooter") %>%
  select(-n) %>%
  group_by(shooter, scoring_play) %>%
  summarize(count = n()) %>%
  pivot_wider(names_from = scoring_play, values_from = count) %>%
  rename(`shot missed 2` = `FALSE`, `shot made 2` = `TRUE`) %>%
  mutate(`non clutch shot percentage` = `shot made 2` /(`shot made 2`+`shot missed 2`)) %>%
  arrange(desc(`non clutch shot percentage`)))
```


Join these tables and create a plot for all players with at least 15 shots in clutch situations that compares their clutch and non-clutch shooting percentage. 
```{r}
non_clutch_situations %>%
  inner_join(clutch_players, by = "shooter") %>%
  mutate(max_percent = case_when(`clutch shot percentage` > `non clutch shot percentage` ~ `clutch shot percentage`, 
                                TRUE ~ `non clutch shot percentage`)) %>%
   rename(`non clutch` = `non clutch shot percentage`, `clutch` = `clutch shot percentage`) %>%
  pivot_longer(cols = c(`non clutch`, `clutch`), names_to = "clutch situation?", values_to = "shot percentage") %>%
  ggplot(aes(y=fct_reorder(shooter, `max_percent`), x=`shot percentage`, fill = `clutch situation?`)) +
  geom_col(position = "identity", alpha = 0.5) + 
  geom_vline(xintercept = 0.5050, color = "darkblue", type = "dashed")+
  labs(y = "", x = "shooting percentage", title = "Who should be shooting in the clutch in the WNBA?", subtitle = "Shooting percentage for players with 15+ shots in Q4 or OT when the score is within one possession", fill = "situation") + 
  scale_fill_manual(values=c("deepskyblue1", "red"))+
  theme_fivethirtyeight() +
  theme(plot.title.position = "plot", 
        plot.title = element_text(size = 13), 
        plot.subtitle = element_text(size = 10))+
  annotate("text", y=1.2, x=0.575, label= "WNBA average \nshooting percentage", cex = 2.5, fontface = "bold", color = "darkblue") +
  scale_x_continuous(expand = c(0,0)) 
```

### Visualization 2 

Create a visualization that shows the breakdown on shooting percentage in clutch versus non-clutch situations by shot type. I broke down the shot types into layups, jump shots, free throws, hook shots, tip shots, and dunks though only displayed layups, jump shots, and free throws because these shots had by far the most data. 
```{r}
(shot_types <- wnba_shots_data %>%
  mutate(clutch_time = period_display_value %in% c("4th Quarter", "OT") & abs(home_score - away_score) <= 3) %>%
  select(clutch_time, period_display_value, home_score, `away_score`, everything()) %>%
  mutate(shot_group = case_when(
    type_text %in% c("Alley Oop Layup Shot", "Cutting Finger Roll Layup Shot", "Cutting Layup Shot", "Driving Finger Roll Layup", "Driving Layup Shot", "Finger Roll Layup", "Layup Driving Reverse", "Layup Running Reverse", "Layup Shot", "Layup Shot Putback", "Reverse Layup Shot", "Running Finger Roll Layup", "Running Layup Shot") ~ "Layup", 
    type_text %in% c("Cutting Dunk Shot", "Driving Dunk Shot", "Reverse Dunk Shot", "Running Dunk Shot") ~ "Dunk",
    type_text %in% c("Driving Floating Bank Jump Shot", "Driving Floating Jump Shot", "Driving Jump Shot Bank", "Fade Away Bank Jump Shot", "Fade Away Jump Shot", "Floating Jump Shot", "Jump Shot", "Jump Shot Bank", "Pullup Bank Jump Shot", "Pullup Jump Shot", "Putback Dunk Shot", "Running Jump Shot", "Running Pullup Jump Shot", "Step Back Bank Jump Shot", "Step Back Jump Shot", "Turnaround Bank Jump Shot", "Turnaround Fade Away Jump Shot", "Turnaround Fadeaway Bank Jump Shot", "Turnaround Jump Shot") ~ "Jump Shot", 
    type_text %in% c("Driving Hook Shot", "Hook Driving Bank", "Hook Shot", "Hook Shot Bank", "Hook Turnaround Bank", "Turnaround Hook Shot") ~ "Hook Shot", 
    type_text %in% c("Free Throw - 1 of 1", "Free Throw - 1 of 2", "Free Throw - 1 of 3", "Free Throw - 2 of 2", "Free Throw - 2 of 3", "Free Throw - 3 of 3", "Free Throw - Clear Path 1 of 2", "Free Throw - Clear Path 2 of 2", "Free Throw - Flagrant 1 of 1", "Free Throw - Flagrant 1 of 2", "Free Throw - Flagrant 1 of 3", "Free Throw - Flagrant 2 of 2", "Free Throw - Flagrant 2 of 3", "Free Throw - Flagrant 3 of 3", "Free Throw - Technical") ~ "Free Throw",
    type_text == "Tip Shot" ~ "Tip Shot", 
    TRUE ~ type_text)) %>%
  group_by(shot_group, scoring_play, clutch_time) %>%
  count())

(shot_percentage_grouped <- tibble(shot_type = c("Free Throw", "Free Throw", "Jump Shot", "Jump Shot", "Layup", "Layup"), percentage = c(1674/2078, 122/159, 111/324, 1868/5145, 92/179, 1390/2497), clutch_time = c(FALSE, TRUE, TRUE, FALSE, TRUE, FALSE)))
  
```

Create the plot 
```{r}
shot_percentage_grouped %>%
  ggplot(aes(y=shot_type, x=percentage, fill = clutch_time)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("dodgerblue3", "darkred"))+
  labs(title = "WNBA shooting percentage drops under stress", y = "", x = "", fill = "Clutch situation?", subtitle = "<strong><span style='color:#8D0000'>Clutch situations</span></strong></b> are defined as shots taken in a one possession game in Q4 or OT") + 
  theme_fivethirtyeight() + 
  theme(plot.title.position = "plot", 
        plot.subtitle = element_markdown(), 
        legend.position = "none")+
  scale_x_continuous(expand = c(0,0)) + 
  geom_curve(
  aes(x = 0.67, y = 2.6, xend = 0.55, yend = 3.25),
  data = shot_percentage_grouped,
  arrow = arrow(length = unit(0.03, "npc"))
)+
  annotate(geom = "text", x = 0.67, y=2.35, label = "Layups shot in the clutch \nhave a 4.27% lower shot accuracy \nthan non-clutch situations", fontface="bold", cex = 3)
```

### Visualization 3
Find average shooting percentage in the WNBA: 
```{r}
wnba_shots_data %>%
  group_by(scoring_play) %>%
  count()
```

The average WNBA shooting percentage is 0.5050. 

Goal is to display shots taken clutch situations for shooters with an above average shooting percentage in the clutch versus those with a below average shooting percentage in the clutch. Data is taken from all players with at least 15 shots clutch situations. 

```{r}
test1 <- wnba_shots_data %>%
  mutate(shooter = word(text, 1, 2)) %>%
  filter(
    shooter %in% c("Jonquel Jones", "Jewell Loyd", "Marina Mabrey", "Skylar Diggins-Smith", "Brittney Griner", "DeWanna Bonner", "Diana Taurasi", "Diamond DeShields", "Breanna Stewart", "Courtney Vandersloot", "Courtney Williams"), 
    coordinate_x > 0, coordinate_y >0, coordinate_y <35,
    period_display_value %in% c("4th Quarter", "OT"), abs(home_score - away_score) <= 3) %>%
  mutate(above_average_shooter = shooter %in% c("Jonquel Jones", "Jewell Loyd", "Marina Mabrey", "Skylar Diggins-Smith", "Brittney Griner")) 

# Make the same plot as above but facet by the team_name to display
# each team's shot charts separately
geom_basketball(league = "WNBA", full_surf = FALSE, rotate = TRUE) +
  geom_point(data = test1, 
             aes(x = coordinate_x - 25, y = coordinate_y - 47 +4,
                 color = scoring_play),
             alpha = 0.8, size = 1) + # modifying the point size
  scale_color_manual(values = c("firebrick1", "cyan")) +
  # Facet by team name and make 4 columns of plots
  facet_wrap(~ above_average_shooter)
```


