---
title: "Exploratory Data Analysis on Shot Data for the 2021 WNBA season"
subtitle: "Nick Ho, Matthew Yep, Erin Franke"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  fig.show = TRUE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#1381B0",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

```{r, include = FALSE, warning = FALSE}

## loading in the data
library(wehoop)
library(tidyverse)
library(lubridate)
library(ggthemes)
```


## What are the distance shooting patterns for the seattle team over time?

```{r, include = FALSE}


## Preparing the data ----------------
sea_wgames <- readRDS("nick_data/sea_wgames.rds")


```

---
# A way to quantify the shooting distance as a trajectory


```{r, echo=FALSE}
library(readr)


sea_df <- read_csv("nick_data/sea_df.csv")
sea_df <- sea_df %>% 
  mutate(game_id = as.character(game_id),
         min_idx = as.integer(min_idx)) %>%
  mutate(game_date = case_when(game_id == 401320570 ~ "LV @ SEA 5/15", 
          game_id == 401320578 ~ "SEA @ MN 5/20", 
          game_id == 401320583 ~ "SEA @ DAL 5/22", 
          game_id == 401320590 ~ "CONN @ SEA 5/25", 
          game_id == 401320595 ~ "MN @ SEA 5/28", 
          game_id == 401320605 ~ "IND @ SEA 6/1", 
          game_id == 401320610 ~ "DAL @ SEA 6/4", 
          game_id == 401320617 ~ "SEA @ ATL 6/9", 
          TRUE ~ game_id))

sea_df %>% 
  ggplot(aes(x = min_idx, y = sea_traj, color = game_date)) +
  geom_line() + 
  geom_point() + 
  ylim(0,60) + 
  labs(title = "The SEA Distance Trajectories for 8 games binned per each 5 minutes", x = "minute index (5 minute intervals)", y = "Seattle trajectory", color = "game") + 
  theme_minimal()+
  theme(plot.title = element_text(size = 15), 
        plot.background = element_rect("aliceblue"), 
        axis.text.y = element_text(size = 15)) 

```

---
# dist matrix idk

```{r, include=FALSE}
total_games <- c(401320570, 401320578, 401320583, 401320590, 401320595, 401320605, 401320610, 401320617)

game_dist <- readRDS("nick_data/game_dist.RData")
game_hclust <- hclust(game_dist, method = "complete")

## assign the lookup table based on how many clusters we want
game_cluster_lookup <- tibble(game_id = as.factor(total_games) , game_cluster = as.factor(cutree(game_hclust, k = 3)))


## annotating the original distance_percent_df with the clusters assigned
lookup_table <- sea_wgames %>% 
  mutate(game_id = as.factor(game_id)) %>% 
  dplyr::select(game_id, game_title) %>% 
  right_join(game_cluster_lookup, by = c("game_id" = "game_id")) %>% 
  unique()

lookup_table[5,2] = "SEA (W) vs MIN (L) pt2"
lookup_table[7,2] = "SEA (W) vs DAL (L) pt2"

## annotating the original distance_percent_df with the clusters assigned
annotated_game_df <- sea_df %>% 
  left_join(lookup_table, by = c("game_id" = "game_id"))

```


.pull-left[
```{r, echo=FALSE, fig.height=4}
plot(game_hclust)
```
]

.pull-right[
```{r, echo=FALSE, fig.height=8}
annotated_game_df %>% 
  ggplot() + 
  geom_line(aes(x = min_idx, y = sea_traj, color = game_title, group = game_id)) +
  facet_wrap(~game_cluster, ncol = 1) + 
  # scale_x_continuous(breaks = seq(0, )) +
  labs(title = "Binned (summed) shot distance for each team binned per distance", x = "Feet from Hoop", y = "Summed Distance", color = "game") + 
  theme_minimal()+
  theme(plot.title = element_text(size = 15))
```
]

---
# Interpreting the grouped trajectories


.pull-left[
```{r, fig.height=7, echo=FALSE}
sea_wgames %>% 
  filter(distance <= 30 & period_number < 5) %>% 
  filter(game_cluster == 1) %>%
  filter(wl_abbrv == "SEA (W)") %>% 
  ggplot() + 
  geom_vline(xintercept = 600, , color = "blue", size=1.5, alpha = 0.3) +
  geom_vline(xintercept = 600*2, , color = "blue", size=1.5, alpha = 0.3)+
  geom_vline(xintercept = 600*3, , color = "blue", size=1.5, alpha = 0.3) +
  geom_point(aes(x = time_corrected, y = distance, color = shot_type, size = 0.5)) +
  scale_shape_manual(values=c(4, 19)) +
  geom_density2d(aes(x = time_corrected, y = distance)) +
  # facet_wrap(~ game_cluster, ncol =1) + 
  # scale_color_manual(values=c("cadetblue", "firebrick")) +
  scale_x_time() +
  labs(title = "Development of distance through the 4 quarters for CLUSTER 1", y = "Distance From Hoop", x = "Time") + 
  theme_bw()
```

]

.pull-right[
```{r, fig.height=7, echo=FALSE}
sea_wgames %>% 
  filter(distance <= 30 & period_number < 5) %>% 
  filter(game_cluster == 2) %>%
  filter(wl_abbrv == "SEA (W)") %>% 
  ggplot() + 
  geom_vline(xintercept = 600, , color = "blue", size=1.5, alpha = 0.3) +
  geom_vline(xintercept = 600*2, , color = "blue", size=1.5, alpha = 0.3)+
  geom_vline(xintercept = 600*3, , color = "blue", size=1.5, alpha = 0.3) +
  geom_point(aes(x = time_corrected, y = distance, color = shot_type, size = 0.5)) +
  scale_shape_manual(values=c(4, 19)) +
  geom_density2d(aes(x = time_corrected, y = distance)) +
  # facet_wrap(~ game_cluster, ncol =1) + 
  # scale_color_manual(values=c("cadetblue", "firebrick")) +
  scale_x_time() +
  labs(title = "Development of distance through the 4 quarters for CLUSTER 2", y = "Distance From Hoop", x = "Time") + 
  theme_bw()
```
]

---

## The Seattle team game + shot movements can be primarily put into 2 categories

.pull-left[
- Cluster 1 attempts more shots in the paint (1st Quarter) 
- Cluster 2 attempts much more shots in the mid range (2nd Quarter)
- Both Cluster 1 and Cluster 2 don't use the mid-range in the 3rd and 4th Quarter
]

.pull-right[
```{r, echo=FALSE, fig.height = 6, fig.width=9}
bin_counts <- readRDS("nick_data/SEA_bin_counts.rds")

bin_counts %>% 
  filter(game_cluster %in% c(1, 2)) %>%
  ggplot(aes(x = shot_type, y = frac_shot_by_quarter)) + 
  facet_wrap(~period_number ) +
  geom_bar(aes(fill = game_cluster) , stat="identity", position = "dodge") +
  labs(title = "Type of Shots taken by each clustered team Normalized by the quarter") +
  theme_bw()
```
]





