---
title: "Exploratory Data Analysis on Shot Data for the 2021 WNBA season"
subtitle: "Nick Ho, Matthew Yep, Erin Franke"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  fig.show = TRUE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#1381B0",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

```{r, include = FALSE, warning = FALSE}

## loading in the data
library(wehoop)
library(tidyverse)
library(lubridate)

library(ggthemes)
library(ggdendro)
library(protoclust)


```


## Question 3: How does the places SEA shoot from change over time?

- Are there any patterns or common groupings we can find in SEA's winning games?


```{r, include = FALSE}


## Preparing the data ----------------
sea_wgames <- readRDS("nick_data/sea_wgames.rds")


```

---
# A way to quantify the shooting distance as a trajectory


```{r, echo=FALSE}
library(readr)


sea_df <- read_csv("nick_data/sea_df.csv")
sea_df <- sea_df %>% 
  mutate(game_id = as.character(game_id),
         min_idx = as.integer(min_idx)) %>%
  mutate(game_date = case_when(game_id == 401320570 ~ "SEA vs LV 5/15", 
          game_id == 401320578 ~ "SEA vs MN 5/20", 
          game_id == 401320583 ~ "SEA vs DAL 5/22", 
          game_id == 401320590 ~ "SEA vs CONN 5/25", 
          game_id == 401320595 ~ "SEA vs MN 5/28", 
          game_id == 401320605 ~ "SEA vs IND 6/1", 
          game_id == 401320610 ~ "SEA vs DAL 6/4", 
          game_id == 401320617 ~ "SEA vs ATL 6/9", 
          TRUE ~ game_id))

sea_df %>% 
  ggplot(aes(x = min_idx, y = sea_traj, color = game_date)) +
  geom_line() + 
  geom_point() + 
  ylim(0,60) + 
  labs(title = "The SEA Distance Trajectories for 8 games binned per each 5 minutes", x = "minute index (5 minute intervals)", y = "Seattle (Summed) Trajectory", color = "game") + 
  theme_minimal()+
  theme(plot.title = element_text(size = 15), 
        plot.background = element_rect("aliceblue"), 
        axis.text.y = element_text(size = 15)) 

```

---
## Distance Between Team-Trajectories

- (simple solution) Euclidian distance between trajectories (A, B)
  - Add the difference between each bucket for A and B and square root the result
  
- Hierarchical clutering on the distance matrix

```{r, include=FALSE}
total_games <- c(401320570, 401320578, 401320583, 401320590, 401320595, 401320605, 401320610, 401320617)

game_dist <- readRDS("nick_data/game_dist.RData")
game_hclust <- hclust(game_dist, method = "complete")

## assign the lookup table based on how many clusters we want
game_cluster_lookup <- tibble(game_id = as.factor(total_games) , game_cluster = as.factor(cutree(game_hclust, k = 3)))


## annotating the original distance_percent_df with the clusters assigned
lookup_table <- sea_wgames %>% 
  mutate(game_id = as.factor(game_id)) %>% 
  dplyr::select(game_id, game_title) %>% 
  right_join(game_cluster_lookup, by = c("game_id" = "game_id")) %>% 
  unique()

lookup_table[5,2] = "SEA (W) vs MIN (L) pt2"
lookup_table[7,2] = "SEA (W) vs DAL (L) pt2"

## annotating the original distance_percent_df with the clusters assigned
annotated_game_df <- sea_df %>% 
  left_join(lookup_table, by = c("game_id" = "game_id"))

```

---

# Clustering Team-Trajectories

.pull-left[
```{r, echo=FALSE, fig.height=6, fig.width=8}
game_hclust$labels <- lookup_table$game_title
ggdendrogram(game_hclust, size = 2) + labs(y = "Dissimilarity between games", title = "Cluster Dendrogram of the SEA games")  +
  theme(plot.title = element_text(size = 14, face = "bold"), axis.text.x = element_text(size = 14, face = "bold")) 
```
]


.pull-right[
```{r, echo=FALSE, fig.height=8, fig.width=8}
annotated_game_df %>% 
  ggplot() + 
  geom_line(aes(x = min_idx, y = sea_traj, color = game_title, group = game_id)) +
  facet_wrap(~game_cluster, ncol = 1) + 
  # scale_x_continuous(breaks = seq(0, )) +
  labs(title = "Binned (summed) shot distance for each team binned per distance", x = "Feet from Hoop", y = "Summed Distance", color = "game") + 
  theme_minimal()+
  theme(plot.title = element_text(size = 15))
```
]

---
## SEA shows two main shooting tendencies


.pull-left[
```{r, fig.height=8, fig.width=8, echo=FALSE}
sea_wgames %>% 
  filter(distance <= 30 & period_number < 5) %>% 
  filter(game_cluster == 1) %>%
  filter(wl_abbrv == "SEA (W)") %>% 
  ggplot() + 
  geom_vline(xintercept = 600, , color = "blue", size=1.5, alpha = 0.3) +
  geom_vline(xintercept = 600*2, , color = "blue", size=1.5, alpha = 0.3)+
  geom_vline(xintercept = 600*3, , color = "blue", size=1.5, alpha = 0.3) +
  geom_point(aes(x = time_corrected, y = distance, color = shot_type, size = 0.5)) +
  scale_shape_manual(values=c(4, 19)) +
  geom_density2d(aes(x = time_corrected, y = distance)) +
  scale_x_time() +
  labs(title = "Shot distance from games in Cluster 1 shows no \n midrange and later 3pt focus in 4th Quarter", y = "Distance From Hoop", x = "Time") + 
  theme_bw()+
  guides(size = FALSE) + 
  theme(plot.title = element_text(size = 20,face = "bold"), 
        axis.text.x = element_text(size = 14, face = "bold"),
        axis.text.y = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 20, face = "bold"),
        legend.key.size = unit(2, 'cm'),
        legend.title  = element_text(size=15),
        legend.text  = element_text(size=15),
        ) 
```

]

.pull-right[
```{r, fig.height=8, fig.width=8, echo=FALSE}
sea_wgames %>% 
  filter(distance <= 30 & period_number < 5) %>% 
  filter(game_cluster == 2) %>%
  filter(wl_abbrv == "SEA (W)") %>% 
  ggplot() + 
  geom_vline(xintercept = 600, , color = "blue", size=1.5, alpha = 0.3) +
  geom_vline(xintercept = 600*2, , color = "blue", size=1.5, alpha = 0.3)+
  geom_vline(xintercept = 600*3, , color = "blue", size=1.5, alpha = 0.3) +
  geom_point(aes(x = time_corrected, y = distance, color = shot_type, size = 0.5)) +
  scale_shape_manual(values=c(4, 19)) +
  geom_density2d(aes(x = time_corrected, y = distance)) +
  # facet_wrap(~ game_cluster, ncol =1) + 
  # scale_color_manual(values=c("cadetblue", "firebrick")) +
  scale_x_time() +
  labs(title = "Shot distance from games in Cluster 2 show more \n midrange and earlier 3pt focus in 4th Quarter", y = "Distance From Hoop", x = "Time") + 
  theme_bw()+
  guides(size = FALSE) + 
  theme(plot.title = element_text(size = 20,face = "bold"), 
        axis.text.x = element_text(size = 14, face = "bold"),
        axis.text.y = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 20, face = "bold"),
        legend.key.size = unit(2, 'cm'),
        legend.title  = element_text(size=15),
        legend.text  = element_text(size=15),
        ) 
```
]


---

# Clustering teams by shot accuracy?

```{r, include=FALSE}

wsd <- readRDS("nick_data/wsdHclust.rds")
distance_percent_df <- readRDS("nick_data/distance_percent_df.rds")
team_abbrv_list <- unique(wsd$away_team_abbrev)


```


```{r, echo=FALSE}
distance_percent_df <- distance_percent_df %>% 
  mutate(feet_from_hoop = idx * 2)

distance_percent_df %>% 
  ggplot() + 
  geom_line(aes(x = feet_from_hoop, y = percent, color = team_id)) +
  labs(title = "Binned Shooting Percentage for each team binned per 2 feet", x = "Feet from Hoop", y = "Shooting Percentage") + 
  theme_bw() + 
  labs(color='Teams') 
```

---
# Top teams show similar accuracy patterns

--
.pull-left[
- Weighted Euclidean Distance


- $$ dist(a, b) = \sqrt{ \sum{} w_i (a_i - b_i)^2}$$

- Weighted by where shots are most common

```{r, echo=FALSE}
bin_weights_df <- readRDS("nick_data/bin_weights_df.rds")

bin_weights_df %>% 
  ggplot(aes(y = bin_weights, x = feet)) +
  geom_line() + 
  geom_point(aes(size = 3, color="red")) + 
  guides(color = FALSE, size = FALSE) + 
  labs(title = "Layups and 3pt bins have the largest weights", x = "Feet from Hoop", y = "Bin Weights") + 
  theme_minimal()

```


]

--

.pull-right[

```{r, echo=FALSE}

team_hclust <- readRDS("nick_data/team_hclust.rds")
# plot(team_hclust)

## assign the lookup table based on how many clusters we want
team_cluster_lookup <- tibble(team_names = team_abbrv_list , team_cluster = as.factor(cutree(team_hclust, k = 3)))

## annotating the original distance_percent_df with the clusters assigned
annotated_distance_percent_df <- distance_percent_df %>% 
  left_join(team_cluster_lookup, by = c("team_id" = "team_names"))

annotated_distance_percent_df <- annotated_distance_percent_df %>% 
  mutate(feet_from_hoop = idx * 2)

ggdendrogram(team_hclust, size = 2) + labs(title = "Top teams share similar scoring accuracy patterns", y = "Dissimilarity of Team Score Accuracies")  +
  theme(plot.title = element_text(size = 14, face = "bold"), axis.text.x = element_text(size = 20, face = "bold"))

# plot(team_hclust)

```

```{r, echo=FALSE, fig.height = 5}

# ## graph the clusters
annotated_distance_percent_df %>%
  ggplot() +
  geom_line(aes(x = feet_from_hoop, y = percent, color = team_id, group = team_id)) +
  # scale_x_continuous(breaks = seq(0, )) +
  facet_wrap(~team_cluster, ncol=1) +
  labs(title = "Binned Shooting Percentage for each team binned per distance", x = "Feet from Hoop", y = "Shooting Percentage") +
  theme_bw() +
  theme(plot.title = element_text(size = 20, face = "bold"), 
        axis.title = element_text(size = 15, face = "bold"),)

```
]


---

# Top teams also show similar defensive patterns

--

.pull-left[

- Defense: How much a team decreases the shot accuracy of another team in a game
- $$D_x = \frac{\sum^n_1 (a_x - b_x)}{n}$$
  - D is the defense for a team for that bin
  - a is the average shooting of opponent over all games (previous slide)
  - b is the average shooting score for opponent in current
  - x is the current bucket (2 feet)
]

--

.pull-right[

```{r, echo = FALSE}

def_percent_df <- readRDS("nick_data/def_percent_df.rds")

def_percent_df %>% 
  ggplot() + 
  geom_line(aes(x = idx, y = def_percent, color = team_id)) +
  labs(title = "Binned Defensive Percentage for each team binned per distance", x = "Bucket Feet from Hoop", y = "Defense_Percentage") + 
  theme_bw()

def_team_hclust <- readRDS("nick_data/def_team_hclust.rds")


ggdendrogram(def_team_hclust, size = 2) + labs(title = "Top teams share similar defensive patterns", y = "Dissimilarity of Team Defensiveness")  +
  theme(plot.title = element_text(size = 14, face = "bold"), axis.text.x = element_text(size = 20, face = "bold"))


```
]




