---
title: "WNBA EDA Project"
subtitle: "\"lets get this bread\" - Yoda"  
author: 
  - "Erin Franke"
  - "Matthew Yep"
  - "Nick Ho"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  fig.show = TRUE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "firebrick",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

## Introduction

Text can be **bold**, _italic_, ~~strikethrough~~, or `inline code`.

[Link to another slide](#colors).

### Lorem Ipsum

Dolor imperdiet nostra sapien scelerisque praesent curae metus facilisis dignissim tortor. 
Lacinia neque mollis nascetur neque urna velit bibendum. 
Himenaeos suspendisse leo varius mus risus sagittis aliquet venenatis duis nec.

- Dolor cubilia nostra nunc sodales

- Consectetur aliquet mauris blandit

- Ipsum dis nec porttitor urna sed

---
name: Data Structure

## How the data is structured

Each row represents a shot taken by a WNBA player


There are many many columns, but we are mainly interested in the coordinates of the shot locations, shots taken in the 4th quarter, and how shots differed across quarters


<!-- New Slide -->
---
# Question 1

Do the best teams in the WNBA take different kinds of shots than the rest of the league?

--
### sub point 1


<!-- New Slide -->
---
class: inverse center middle
# Question 2

--
```
class: inverse center middle
```

<!-- New Slide -->
---
# Question 3

```{r, echo=FALSE}

a <- 2
print("the code for this chunk will not be displayed")

```

Nick is edititing this line. Did I spell it correctly? Who knows :)))

<!-- New Slide -->
---
# Question 4

I might have just been dumb about this but this is a quick
test about how this might work

This is an additional edit on my end


