---
title: " Analysis on SEA Team Shooting Distance Trajectories"
author: "Nick"
date: "6/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Here is the code to my slides where I explored the question **How do the places SEA shoot from change over time?**

## Loading in the data and initial wrangling


Grabbing all the games that seattle time won. (I got the game_ids from a different script that looped through all games to count which team one.) 

```{r, message=FALSE, warning=FALSE}
library(wehoop)
library(tidyverse)
library(lubridate)
library(ggdendro)
## Preparing the data ----------------


wnba_pbp_data <- load_wnba_pbp(2021)

# Get the shots and clean this data a bit:
wnba_shots_data <- wnba_pbp_data %>%
  filter(shooting_play)
# Remove unnecessary columns:
wnba_shots_data <- wnba_shots_data %>%
  dplyr::select(-shooting_play, -id, -participants_2_athlete_id,
                -type_abbreviation, -season, -season_type, 
                -home_team_spread, -game_spread, -home_favorite)

write.csv(wnba_shots_data,'WNBA_shot_data.csv')
## Loading the data for a single game ----------------

wsd <-wnba_shots_data

list_of_SEA_won_games <- c(401320570, 401320578, 401320583, 401320590, 401320595, 401320605, 401320610, 401320617) # removed: none
list_of_games <- unique(wsd$game_id)

##### looping through the 8 games and added them up together

# initializing empty dataframe
sea_wgames = data.frame()

for (curr_game in list_of_SEA_won_games) {
  print(paste("=> curr game: ", curr_game))


single_game <- wsd %>% 
  filter(game_id == curr_game)
#game 1: 401320570 401320578 401320583 401320590 401320595 401320605 401320610 401320617

sea_games <- wsd %>% 
  filter()

## Loading the data for a single game ----------------

sg <- single_game %>% 
  mutate(distance = sqrt((coordinate_x-25)^2 + coordinate_y^2)) %>% 
  filter(coordinate_x > 0 & coordinate_y > 0) %>% 
  mutate(sequence_number = as.integer(sequence_number))


## Add the time (min:sec) column to the game data ----------------

sg <- sg %>% 
  # select(clock_minutes, clock_seconds, period_number, away_score, home_score) %>% 
  mutate(seconds_elapsed_game = 600 - (as.integer(clock_minutes) * 60 + as.integer(clock_seconds))) %>% 
  mutate(seconds_with_period = (period_number - 1) * 10 * 60 + seconds_elapsed_game) %>% 
  mutate(time_corrected = seconds_to_period(seconds_with_period))



## Adding the winning team identifier column ----------------

final_score_home <- sg$home_score[length(sg$home_score)]
# print(final_score_home)
final_score_away <- sg$away_score[length(sg$away_score)]
# print(final_score_away)

away_id <- sg$away_team_id[1]
home_id <- sg$home_team_id[1]

away_abbrv <- sg$away_team_abbrev[1]
home_abbrv <- sg$home_team_abbrev[1]

if ( final_score_home > final_score_away) {
  winning_id = sg$home_team_id[1]
  losing_id = sg$away_team_id[1]
  win_abbrv = home_abbrv
  lose_abbrv = away_abbrv
} else {
  winning_id = sg$away_team_id[1]
  losing_id = sg$home_team_id[1]
  win_abbrv = away_abbrv
  lose_abbrv = home_abbrv
}

sg <- sg %>% 
  mutate(wl_abbrv = case_when(team_id == winning_id ~ paste(win_abbrv, "(W)"),
                              team_id == losing_id ~ paste(lose_abbrv, "(L)")),
         game_title = paste(paste(win_abbrv, "(W) vs",paste(lose_abbrv, "(L)"))))

print(paste(paste(win_abbrv, "(W) vs",paste(lose_abbrv, "(L)"))))

sea_wgames = rbind(sea_wgames, sg)
}

```

## Creating the trajectories

I'll create the trajectories by going over bins of 2 feet, distance from the
hoop. I wrote this before I was aware of the cut function (which I use in my other scripts).


```{r, warning=FALSE}


### grabbing and graphing the delta distance trajectories so that I can cluster them

total_games <- c(401320570, 401320578, 401320583, 401320590, 401320595, 401320605, 401320610, 401320617)
games_of_interest <-  c(401320570)

distance_df <- data.frame()
sea_df <- data.frame()
other_df <- data.frame()

for ( curr_game_id in total_games){
  # print(curr_game_id)
  
  curr_game <- sea_wgames %>% 
    filter(game_id == curr_game_id)
  
  sea_traj = c()
  other_traj = c()
  delta_distance_per_window = c()

  for ( i in seq(1,40, 5)){
    
    # print(i)
    start_sec <- 60 * (i - 1)
    end_sec <-  60 * (i)
    
    temp <- curr_game %>%
      filter(time_corrected > start_sec & time_corrected < end_sec) 
    
    win_distance_hit = sum(filter(temp, (wl_abbrv == "SEA (W)") & scoring_play == TRUE)$distance)
    win_distance_miss = sum(filter(temp, (wl_abbrv == "SEA (W)") & scoring_play == FALSE)$distance)
    
    other_distance_hit = sum(filter(temp, (wl_abbrv != "SEA (W)") & scoring_play == TRUE)$distance)
    other_distance_miss = sum(filter(temp, (wl_abbrv != "SEA (W)") & scoring_play == FALSE)$distance)
    
    if(is.na(win_distance_hit)){win_distance <-  0}
    if(is.na(win_distance_miss)){other_distance <- 0}
    if(is.na(other_distance_hit)){win_distance <-  0}
    if(is.na(other_distance_miss)){other_distance <- 0}
    
    win_distance = win_distance_hit * 1 + win_distance_miss
    other_distance = other_distance_hit * 1 + other_distance_miss
    
    dist_delta <- win_distance - other_distance
    
    delta_distance_per_window = c(delta_distance_per_window, dist_delta)
    
    ## just grabbing the seattle distance evolution 
    sea_traj = c(sea_traj, sum(filter(temp, (wl_abbrv == "SEA (W)"))$distance)) ## alternate between mean and sum
    other_traj = c(other_traj, sum(filter(temp, (wl_abbrv != "SEA (W)"))$distance)) ## alternate between mean and sum
    
  }
  min_df = data.frame(min_idx = seq(1, length(delta_distance_per_window)), delta_distance_per_window, game_id = as.factor(curr_game_id))
  sea_min_df = data.frame(min_idx = seq(1, length(delta_distance_per_window)), sea_traj, other_traj, game_id = as.factor(curr_game_id))
  other_min_df = data.frame(min_idx = seq(1, length(delta_distance_per_window)), other_traj, game_id = as.factor(curr_game_id))
  
  distance_df <- rbind(distance_df , min_df)
  sea_df <- rbind(sea_df , sea_min_df)
  other_df <- rbind(other_df , other_min_df)
  
}

sea_df %>% 
  ggplot(aes(x = min_idx, y = sea_traj, color = game_id)) +
  geom_line() + 
  geom_point() + 
  ylim(0,60) + 
  labs(title = "The SEA Distance Trajectories for 8 games binned per each 5 minutes")
```

## Pairwise distance function


```{r}

calc_traj_pairwise <- function(df, game_id1, game_id2){
  
  num_bins = length(unique(df$min_idx))
  total_distance <- 0
  
  #pairwise distance between each datapoint and adding them together
  for(i in seq(num_bins)){
    game1_d <- df %>% 
      filter(game_id == game_id1 & min_idx == i)
    
    game2_d <- df %>% 
      filter(game_id == game_id2 & min_idx == i)
    
    bin_euc_distance <- (game1_d$sea_traj - game2_d$sea_traj)^2 ## bin_weight
    
    total_distance = total_distance + bin_euc_distance
  }
  total_distance = sqrt(total_distance)
  
  return (total_distance)
}

## checking
calc_traj_pairwise(sea_df, 401320583, 401320578)

```


## Creating the distance matrix and clustering

```{r}

total_games <- c(401320570, 401320578, 401320583, 401320590, 401320595, 401320605, 401320610, 401320617)


team_pairwise_dist <- map_dfr(total_games, 
        function(game1_id){
          map_dfr(total_games, 
                  function(game2_id){
                    tibble("game1" = game1_id, "game2" = game2_id, 
                           "distance" = calc_traj_pairwise(sea_df, game1_id, game2_id))
          })
        }
      )

## pivot wider
dist_wider <- pivot_wider(team_pairwise_dist, names_from = game1, values_from = distance)

## removing the first label column called team1, then setting it into a matrix
dist_matrix <- dist_wider %>% 
  dplyr::select(-game2) %>% 
  as.matrix()

## setting the row names to be the abbrv list, which is the same order as before
rownames(dist_matrix) <- total_games

## creating a distance object out of it
game_dist <- as.dist(dist_matrix)


```


```{r}

game_hclust <- hclust(game_dist, method = "complete")



## assign the lookup table based on how many clusters we want
game_cluster_lookup <- tibble(game_id = as.factor(total_games) , game_cluster = as.factor(cutree(game_hclust, k = 3)))


## annotating the original distance_percent_df with the clusters assigned
lookup_table <- sea_wgames %>% 
  mutate(game_id = as.factor(game_id)) %>% 
  dplyr::select(game_id, game_title) %>% 
  right_join(game_cluster_lookup, by = c("game_id" = "game_id")) %>% 
  unique()

lookup_table[5,2] = "SEA (W) vs MIN (L) pt2"
lookup_table[7,2] = "SEA (W) vs DAL (L) pt2"


## annotating the original distance_percent_df with the clusters assigned
annotated_game_df <- sea_df %>% 
  left_join(lookup_table, by = c("game_id" = "game_id"))

## graph the clusters wrapped in different graphs
annotated_game_df %>% 
  ggplot() + 
  geom_line(aes(x = min_idx, y = sea_traj, color = game_title, group = game_id)) +
  facet_wrap(~game_cluster, ncol = 1) + 
  # scale_x_continuous(breaks = seq(0, )) +
  labs(title = "Binned (summed) shot distance for each team binned per distance", x = "Feet from Hoop", y = "Summed Distance")
```

## Dendrogram

Here is the dendrogram demonstrated in the power point.

```{r}

game_hclust$labels <- lookup_table$game_title

ggdendrogram(game_hclust, size = 2) + labs(y = "Dissimilarity between games", title = "Cluster Dendrogram of the SEA games")  +
  theme(plot.title = element_text(size = 14, face = "bold"), axis.text.x = element_text(size = 14, face = "bold")) 

```




